---
title: "K20A Woody plant density"
output: html_notebook
---

Install and load R packages
```{r}
#install.packages("packman")
pacman::p_load(tidyverse, plotrix, grid, ggpubr, car, lme4, lmerTest, emmeans, DHARMa, install = TRUE, update = getOption("pac_update"))

```

Prepping dataset for figure creation and statistical analyses.

Read in csv data
```{r}
k20a_woody_density <- read.csv('please read in csv from the write folder', header = TRUE)
```

Dealing with dates
```{r}
k20a_woody_density$Date_sampled <- as.Date(k20a_woody_density$Date_sampled, format = "%m/%d/%Y")
k20a_woody_density$Year <- lubridate::year(k20a_woody_density$Date_sampled)
```

Organizing species data to different functional group levels
```{r}
k20a_woody_density <- k20a_woody_density %>% 
  mutate(Division = if_else(Genus_species == "Juniperus_virginiana", "Gymnosperm", "Angiosperm"),
         Life_form = case_when(Genus_species %in% c("Acer_negundo", "Cercis_canadensis", "Gleditsia_triacanthos", "Juniperus_virginiana", "Ulmus_americana", "Ulmus_pumila") ~ "Tree",
                               Genus_species %in% c("Cornus_drummondii", "Prunus_americana", "Rhus_aromatica", "Rhus_glabra", "Zanthoxylum_americanum") ~ "Shrub",
                               TRUE ~ "Other"),
         Resprout = case_when(Genus_species %in% c("Acer_negundo", "Amorpha_canescens", "Ceanothus_americanus", "Cercis_canadensis", "Cornus_drummondii", "Gleditsia_triacanthos", "Opuntia_macrorhiza", "Prunus_americana", "Rhus_aromatica", "Rhus_glabra", "Rosa_spp", "Rubus_spp", "Symphoricarpos_orbiculatus", "Toxicodendron_radicans", "Ulmus_americana", "Ulmus_pumila", "Zanthoxylum_americanum") ~ "Resprouter",
                              TRUE ~ "Non-resprouter"))
```


```{r}
k20a_woody_density %>% 
  filter(Life_form == "Shrub") %>% 
  group_by(Genus_species) %>% 
  summarise(Count.avg = mean(Count))
  
```

Top 5 shrub species
1. Cornus drummondii (N = 16161)
2. Prunus americana (N = 3094)
3. Rhus glabra (N = 2570)
4. Zanthoxylum americanum (N = 1581)
5. Rhus aromatica (N = 302)



###Density Figures
####Total Shrub Density
```{r}
k20a_transect_summary <- k20a_woody_density %>% 
  filter(Year >= 2019 & Life_form == "Shrub") %>% 
  group_by(Year, Position, Transect) %>% 
  summarise(sum_dens = sum(Count)/(100*0.5))

k20a_position_summary <- k20a_transect_summary %>% 
  group_by(Year, Position) %>% 
  summarise(Density_avg = mean(sum_dens, na.rm = TRUE),
            Density_SE = std.error(sum_dens, na.rm = TRUE))
  

k20a_position_summary$Position <- factor(k20a_position_summary$Position, levels = c("Upland","Slope","Lowland"))  


all_shrubs <- ggplot(k20a_position_summary, aes(x=Position, y=Density_avg, fill=as.factor(Year)))+
  geom_col(position = position_dodge())+
  geom_errorbar(aes(ymin = Density_avg - Density_SE, ymax = Density_avg + Density_SE), position = position_dodge(width = 0.9), width=.2)+
  scale_y_continuous(limits = c(0,21.5),
                     breaks = c(0,4,8,12,16,20),
                     expand = c(0,0))+
  scale_fill_grey()+
  guides(color=guide_legend(title = "Year"))+
  labs(x = "Topographic Position",
       y = bquote('Stem Density ('~ m^{-2}~")"))+
  theme_bw()+
  theme(axis.title = element_text(size = 18, color = "black", face = "bold"),
        axis.text = element_text(size = 16, color = "black", face = "bold"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        #legend.position = c(0.12,0.80),
        legend.position = "NA",
        legend.title = element_text(size = 14),
        legend.text = element_text(size = 12))

ggsave("shrubs_stem_density.png",
       plot = all_shrubs,
       width = 12,
       height = 8,
       units = "in",
       dpi = 300,
       device = "png")
```


####Cornus drummondii density
```{r}
codr_transect_summary <- k20a_woody_density %>% 
  filter(Year >= 2019 & Genus_species == "Cornus_drummondii" ) %>% 
  group_by(Year, Position, Transect) %>% 
  summarise(TDensity_avg = mean(Density_m2, na.rm = TRUE))


codr_position_summary <- codr_transect_summary %>% 
  group_by(Year, Position) %>% 
  summarise(Density_avg = mean(TDensity_avg, na.rm = TRUE),
            Density_SE = std.error(TDensity_avg, na.rm = TRUE))


codr_position_summary$Position <- factor(codr_position_summary$Position, levels = c("Upland","Slope","Lowland"))  

codr <- ggplot(codr_position_summary, aes(x=Position, y=Density_avg, fill=as.factor(Year)))+
  geom_col(position = position_dodge())+
  geom_errorbar(aes(ymin = Density_avg - Density_SE, ymax = Density_avg + Density_SE), position = position_dodge(width = 0.9), width=.2)+
  scale_y_continuous(limits = c(0,16.5),
                     breaks = c(0,3,6,9,12,15),
                     expand = c(0,0))+
  scale_fill_grey()+
  labs(x = "Topographic Position",
       y = bquote('Stem Density ('~ m^{-2}~")"))+
  guides(color=guide_legend(title = "Year"))+
  theme_bw()+
  theme(axis.title = element_text(size = 18, color = "black", face = "bold"),
        axis.text = element_text(size = 16, color = "black", face = "bold"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "NA")


ggsave("codr_stem_density.png",
       plot = codr,
       width = 12,
       height = 8,
       units = "in",
       dpi = 300,
       device = "png")
```


####Prunus americana density
```{r}
pram_transect_summary <- k20a_woody_density %>% 
  filter(Year >= 2019 & Genus_species == "Prunus_americana" ) %>% 
  group_by(Year, Position, Transect) %>% 
  summarise(TDensity_avg = mean(Density_m2, na.rm = TRUE))


pram_position_summary <- pram_transect_summary %>% 
  group_by(Year, Position) %>% 
  summarise(Density_avg = mean(TDensity_avg, na.rm = TRUE),
            Density_SE = std.error(TDensity_avg, na.rm = TRUE))


pram_position_summary$Position <- factor(pram_position_summary$Position, levels = c("Upland","Slope","Lowland"))  


pram <- ggplot(pram_position_summary, aes(x=Position, y=Density_avg, fill=as.factor(Year)))+
  geom_col(position = position_dodge())+
  geom_errorbar(aes(ymin = Density_avg - Density_SE, ymax = Density_avg + Density_SE), position = position_dodge(width = 0.9), width=.2)+
  scale_y_continuous(limits = c(0,10.5),
                     breaks = c(0,2,4,6,8,10),
                     expand = c(0,0))+
  scale_fill_grey()+
  labs(x = "Topographic Position",
       y = bquote('Stem Density ('~ m^{-2}~")"))+
  guides(color=guide_legend(title = "Year"))+
  theme_bw()+
  theme(axis.title = element_text(size = 18, color = "black", face = "bold"),
        axis.text = element_text(size = 16, color = "black", face = "bold"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "NA")

ggsave("pram_stem_density.png",
       plot = pram,
       width = 12,
       height = 8,
       units = "in",
       dpi = 300,
       device = "png")
```


####Rhus aromatica density
```{r}
rhar_transect_summary <- k20a_woody_density %>% 
  filter(Year >= 2019 & Genus_species == "Rhus_aromatica" ) %>% 
  group_by(Year, Position, Transect) %>% 
  summarise(TDensity_avg = mean(Density_m2, na.rm = TRUE))


rhar_position_summary <- rhar_transect_summary %>% 
  group_by(Year, Position) %>% 
  summarise(Density_avg = mean(TDensity_avg, na.rm = TRUE),
            Density_SE = std.error(TDensity_avg, na.rm = TRUE))


rhar_position_summary$Position <- factor(rhar_position_summary$Position, levels = c("Upland","Slope","Lowland"))  


rhar <- ggplot(rhar_position_summary, aes(x=Position, y=Density_avg, fill=as.factor(Year)))+
  geom_col(position = position_dodge())+
  geom_errorbar(aes(ymin = Density_avg - Density_SE, ymax = Density_avg + Density_SE), position = position_dodge(width = 0.9), width=.2)+
  scale_y_continuous(limits = c(0,0.35),
                     breaks = c(0,0.1,0.2,0.3),
                     expand = c(0,0))+
  scale_fill_grey()+
  labs(x = "Topographic Position",
       y = bquote('Stem Density ('~ m^{-2}~")"))+
  guides(color=guide_legend(title = "Year"))+
  theme_bw()+
  theme(axis.title = element_text(size = 18, color = "black", face = "bold"),
        axis.text = element_text(size = 16, color = "black", face = "bold"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "NA")

ggsave("rhar_stem_density.png",
       plot = rhar,
       width = 12,
       height = 8,
       units = "in",
       dpi = 300,
       device = "png")
```


####Rhus glabra density
```{r}
rhgl_transect_summary <- k20a_woody_density %>% 
  filter(Year >= 2019 & Genus_species == "Rhus_glabra" ) %>% 
  group_by(Year, Position, Transect) %>% 
  summarise(TDensity_avg = mean(Density_m2, na.rm = TRUE))


rhgl_position_summary <- rhgl_transect_summary %>% 
  group_by(Year, Position) %>% 
  summarise(Density_avg = mean(TDensity_avg, na.rm = TRUE),
            Density_SE = std.error(TDensity_avg, na.rm = TRUE))


rhgl_position_summary$Position <- factor(rhgl_position_summary$Position, levels = c("Upland","Slope","Lowland"))  


rhgl <- ggplot(rhgl_position_summary, aes(x=Position, y=Density_avg, fill=as.factor(Year)))+
  geom_col(position = position_dodge())+
  geom_errorbar(aes(ymin = Density_avg - Density_SE, ymax = Density_avg + Density_SE), position = position_dodge(width = 0.9), width=.2)+
  scale_y_continuous(limits = c(0,3.5),
                     breaks = c(0,1,2,3),
                     expand = c(0,0))+
  scale_fill_grey()+
  labs(x = "Topographic Position",
       y = bquote('Stem Density ('~ m^{-2}~")"))+
  guides(color=guide_legend(title = "Year"))+
  theme_bw()+
  theme(axis.title = element_text(size = 18, color = "black", face = "bold"),
        axis.text = element_text(size = 16, color = "black", face = "bold"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "NA")

ggsave("rhgl_stem_density.png",
       plot = rhgl,
       width = 12,
       height = 8,
       units = "in",
       dpi = 300,
       device = "png")
```


####Zanthoxylum americanum density
```{r}
zaam_transect_summary <- k20a_woody_density %>% 
  filter(Year >= 2019 & Genus_species == "Zanthoxylum_americanum" ) %>% 
  group_by(Year, Position, Transect) %>% 
  summarise(TDensity_avg = mean(Density_m2, na.rm = TRUE))


zaam_position_summary <- zaam_transect_summary %>% 
  group_by(Year, Position) %>% 
  summarise(Density_avg = mean(TDensity_avg, na.rm = TRUE),
            Density_SE = std.error(TDensity_avg, na.rm = TRUE))


zaam_position_summary$Position <- factor(zaam_position_summary$Position, levels = c("Upland","Slope","Lowland"))  


zaam <- ggplot(zaam_position_summary, aes(x=Position, y=Density_avg, fill=as.factor(Year)))+
 geom_col(position = position_dodge())+
  geom_errorbar(aes(ymin = Density_avg - Density_SE, ymax = Density_avg + Density_SE), position = position_dodge(width = 0.9), width=.2)+
  scale_y_continuous(limits = c(0,4.5),
                     breaks = c(0,1,2,3,4),
                     expand = c(0,0))+
  scale_fill_grey()+
  labs(x = "Topographic Position",
       y = bquote('Stem Density ('~ m^{-2}~")"))+
  guides(color=guide_legend(title = "Year"))+
  theme_bw()+
  theme(axis.title = element_text(size = 18, color = "black", face = "bold"),
        axis.text = element_text(size = 16, color = "black", face = "bold"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "NA")

ggsave("zaam_stem_density.png",
       plot = zaam,
       width = 12,
       height = 8,
       units = "in",
       dpi = 300,
       device = "png")
```


Organize all figures into one image with several panels.
```{r}
stem_density <- ggarrange(all_shrubs, codr, pram, rhar, rhgl, zaam,
          ncol = 2, nrow = 3, align = "hv",
          common.legend = TRUE)


ggsave("stem_density.png",
       plot = stem_density,
       width = 12,
       height = 12,
       units = "in",
       dpi = 300,
       device = "png")
```


##Shrub Density Stats
Using shrub ramet/stem densities the first year following all cutting treatments.
```{r}
k20a_woody_density_2019to2021 <- k20a_woody_density %>% 
  filter(Year >= 2019)
```


Analyze top 5 species together
```{r}
hist(k20a_transect_summary$sum_dens)
hist(log(k20a_transect_summary$sum_dens))#Best transformation

woody_den_lmer <- lmer(log(sum_dens) ~ as.factor(Year) * Position + (1|Transect),
                       data = k20a_transect_summary, REML = TRUE)

#Test model assumptions
woody_lmer_resid <- simulateResiduals(woody_den_lmer)
plot(woody_lmer_resid)

anova(woody_den_lmer, type = 3, ddf = "Kenward-Roger")
print(summary(woody_den_lmer, type=3, test.statistic = "F",ddf = "Kenward-Roger"))
emmeans(woody_den_lmer, pairwise ~ Position|Year, adjust = "sidak", type = "response")
emmeans(woody_den_lmer, pairwise ~ Year|Position, adjust = "sidak", type = "response")

```

Analyze only Cornus drummondii
```{r}
codr <- k20a_woody_density_2019to2021[k20a_woody_density_2019to2021$Genus_species=="Cornus_drummondii",]

hist(codr$Density_m2)
hist(log(codr$Density_m2))
hist(sqrt(codr$Density_m2))#Best transformation

codr_dens_lmer <- lmer(sqrt(Density_m2) ~ as.factor(Year) * Position + (1|Transect),
                      data = codr, REML = TRUE)

#Test model assumptions
codr_dens_resid <- simulateResiduals(codr_dens_lmer)
plot(codr_dens_resid)

anova(codr_dens_lmer, type = 3, ddf = "Kenward-Roger")
print(summary(codr_dens_lmer, type=3, test.statistic = "F",ddf = "Kenward-Roger"))
emmeans(codr_dens_lmer, pairwise ~ Position|Year, adjust = "sidak", type = "response")
emmeans(codr_dens_lmer, pairwise ~ Year|Position, adjust = "sidak", type = "response")

```

Analyze only Prunus americana
```{r}
pram <- k20a_woody_density_2019to2021[k20a_woody_density_2019to2021$Genus_species=="Prunus_americana",]

hist(pram$Density_m2)
hist(log(pram$Density_m2))#best transformation
hist(sqrt(pram$Density_m2))

pram_dens_lmer <- lmer(sqrt(Density_m2) ~ as.factor(Year) * Position + (1|Transect),
                      data = pram, REML = TRUE)

#Test model assumptions
pram_dens_resid <- simulateResiduals(pram_dens_lmer)
plot(pram_dens_resid)

anova(pram_dens_lmer, type = 3, ddf = "Kenward-Roger")
print(summary(pram_dens_lmer, type=3, test.statistic = "F",ddf = "Kenward-Roger"))
```


Analyze only Rhus aromatica
```{r}
rhar <- k20a_woody_density_2019to2021[k20a_woody_density_2019to2021$Genus_species=="Rhus_aromatica",]

hist(rhar$Density_m2)
hist(log(rhar$Density_m2))#best transformation
hist(sqrt(rhar$Density_m2))

rhar_dens_lmer <- lmer(sqrt(Density_m2) ~ as.factor(Year) * Position + (1|Transect),
                      data = rhar, REML=TRUE)

#Test model assumptions
rhar_dens_resid <- simulateResiduals(rhar_dens_lmer)
plot(rhar_dens_resid)

anova(rhar_dens_lmer, type = 3, ddf = "Kenward-Roger")
print(summary(rhar_dens_lmer, type=3, test.statistic = "F",ddf = "Kenward-Roger"))
```


Analyze only Rhus glabra
```{r}

rhgl <- k20a_woody_density_2019to2021[k20a_woody_density_2019to2021$Genus_species=="Rhus_glabra",]

hist(rhgl$Density_m2)
hist(log(rhgl$Density_m2))#best transformation
hist(sqrt(rhgl$Density_m2))

rhgl_dens_lmer <- lmer(sqrt(Density_m2) ~ as.factor(Year) * Position + (1|Transect),
                      data = rhgl, REML=TRUE)

#Test model assumptions
rhgl_dens_resid <- simulateResiduals(rhgl_dens_lmer)
plot(rhgl_dens_resid)

anova(rhgl_dens_lmer, type = 3, ddf = "Kenward-Roger")
print(summary(rhgl_dens_lmer, type=3, test.statistic = "F",ddf = "Kenward-Roger"))
```

Analyze only Zanthoxylum americanum
```{r}
zaam <- k20a_woody_density_2019to2021[k20a_woody_density_2019to2021$Genus_species=="Zanthoxylum_americanum",]

hist(zaam$Density_m2)
hist(log(zaam$Density_m2))#Best transformation
hist(sqrt(zaam$Density_m2))

zaam_dens_lmer <- lmer(sqrt(Density_m2) ~ as.factor(Year) * Position + (1|Transect),
                      data = zaam, REML = TRUE)

#Test model assumptions
zaam_dens_resid <- simulateResiduals(zaam_dens_lmer)
plot(zaam_dens_resid)

anova(zaam_dens_lmer, type = 3, ddf = "Kenward-Roger")
print(summary(zaam_dens_lmer, type=3, test.statistic = "F",ddf = "Kenward-Roger"))
```



####Tree and Shrub cover change using Rangeland Analysis Platform (RAP) data for K20A
```{r}
RAP <- read.csv('Remember to call in the data file from your folder')
```


Plotting tree data from RAP by landscape position
```{r}
RAP_tree <- 
ggplot(data = RAP)+
  geom_rect(aes(xmin = 2017, xmax = 2018.5, ymin = 0, ymax = 50))+
  geom_point(aes(x=year, y=TRE, color=Location), size=2.5)+
  geom_line(aes(x=year, y=TRE, color=Location), linewidth=1.5)+
  scale_x_continuous(breaks = c(1986:2021),
                     limits = c(1986,2021),
                     expand = c(0,0))+
  labs(x = "Year",
       y = "Tree Cover (%)")+
  scale_y_continuous(limits = c(0,50),
                     breaks = c(0,10,20,30,40,50),
                     expand = c(0,0))+
  scale_color_manual(values = c("#1A80D9","#F2AB27","#030A8C"))+
  guides(color=guide_legend(title = "Landscape Position"))+
  #geom_vline(xintercept = 2018, color="black")+
  theme_bw()+
  theme(axis.title = element_text(size = 18, color = "black", face = "bold"),
        axis.text.x = element_text(size = 14, color = "black", face = "bold", angle=45, hjust = 1),
        axis.text.y = element_text(size = 14, color = "black", face = "bold"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = c(0.15,0.85),
        legend.title = element_text(size = 14),
        legend.text = element_text(size = 12))

ggsave("k20a_tree_cover.png",
       plot = RAP_tree,
       height = 8, width = 12, units = "in",
       device = "png",
       dpi = 300)
```


Basic linear regression analysis for RAP tree cover.
```{r}
tree.lm <- lm(TRE~year*Location, RAP)

plot(tree.lm)#data meets assumptions

summary(tree.lm)
```
Releveling linear regression model to look at all variable significance.
```{r}
RAP_re <- RAP %>% 
  mutate(Location = Location %>% 
           fct_relevel("Upland"))

tree.lm.reorder <- lm(TRE~year*Location, data=RAP_re)

plot(tree.lm.reorder)

summary(tree.lm.reorder)
```


